---
title: 'Predictive Machine Learning: Project'
output:
  html_document:
    highlight: zenburn
    theme: spacelab
    toc: yes
date: '2015-02-22'
---

## Executive Summary

A

## 1. Data Set

The data set we are going to use in the analysis (`mtcars`) consists of 32 observations on 11 variables, as explained in [the manual](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).

Packages and options needed for the analysis.

```{r, message = FALSE}
library(caret)
library(dplyr)
library(tidyr)
library(magrittr)
library(randomForest)
library(pander)
library(knitr)

opts_chunk$set(fig.width = 10, fig.height = 7, warning = FALSE, message = FALSE, cache = TRUE)
panderOptions("table.split.table", Inf)
```

```{r}
training.csv <- read.csv("data/pml-training.csv")
testing.csv <- read.csv("data/pml-testing.csv")
```

## 2. Exploratory Data Analysis

The data set is loaded. Since all the variables are numeric, the `am` variable (transmission type) is recoded to a factor with two levels: *automatic* (originally 0) and *manual* (1).

```{r}
training.csv %>% dim %>% pander
training.csv %$% table(classe) %>% pander
```

In the Appendix, a correlation matrix plot is presented, giving a glance at the data.

## 3. Feature Selection

```{r}
training.long <- training.csv %>% tbl_df %>% gather %>% group_by(key) %>%
    summarise(valid = 1 - sum(is.na(value) | value == "") / n())

training.long %$% table(valid) %>% pander
variables <- training.long %>% filter(valid == 1) %$% key %>% .[8:60]
```

## 4. Model Preparation

```{r}
set <- createDataPartition(training.csv$classe, p = 0.7, list = FALSE)

training.set <- training.csv[set, variables]
validation.set <- training.csv[-set, variables]
testing.set <- testing.csv[, variables[-53]]
```

## 5. Model Fitting

```{r}
fit <- randomForest(classe ~ ., training.set)
confusionMatrix(predict(fit, training.set), training.set$classe)
confusionMatrix(predict(fit, validation.set), validation.set$classe)
```

```{r}
varImpPlot(fit)
```

## 6. Model Tuning

```{r}
varImp(fit) %>% add_rownames %>% arrange(-Overall) %>% head(10)

variables.kiss <- varImp(fit) %>% add_rownames %>% top_n(7, Overall) %$% rowname %>% c("classe")

fit.kiss <- randomForest(classe ~ ., training.set[, variables.kiss])
confusionMatrix(predict(fit.kiss, training.set), training.set$classe)
confusionMatrix(predict(fit.kiss, validation.set), validation.set$classe)
testing.pred <- predict(fit.kiss, testing.set)
testing.pred %>% pander
testing.pred.kiss <- predict(fit.kiss, testing.set)
testing.pred.kiss %>% pander

identical(testing.pred, testing.pred.kiss)
```

## 7. Results

```{r}
pml_write_files <- function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("data/problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
    }
}

pml_write_files(testing.pred.kiss)
```
